---
title: "Statistical Tests from 6304 and 6305"
author: "Darren Keeley"
output:
  html_document:
    toc: true
    theme: united
---
<!--
##Table of Contents
1. One sample T test
2. Two sample T test + Tests for normality
3. Paired T test
4. Wilcoxon Rank Sum / Mann Whitney U np test for 2 independent means
5. Wilcoxon Signed Rank np test for paired continuous data
6. One Way ANOVA + HOV Levene Test
7. Kruskal Wallis np 1 way test
8. Chi Square test for independence + Contingency Table and Fisher's Exact Test
9. Chi Square goodness of fit test
10. Linear Contrasts
11. Tukey
12. Two Way ANOVA + Output residuals and test for normality
13. Completely Randomized Block Design (CRBD)
14. Latin Square Design (LSD)
15. ANCOVA
16. Random Effect and Mixed Effect Models + Varcomp
17. Nested Factors
18. Crossover Design
-->
#IMPORTANT! Run the chunk below to get correct Type III SS. Also loads necessary libraries.
```{r, message=FALSE}
# The following is required to calculate Type III SS when running ANOVA.
# See Note 0
# options("contrasts")
options(contrasts=c('contr.sum','contr.poly'))

# car for Anova().
library(tidyverse); library(car)
```

###1. One sample T test, Week 5 lab
```{r}
paper <- read.csv("data/paper.csv")[2]
```

```{r}
#H0: mean <= 1600
#Ha: mean > 1600

t.test(paper, mu=1600, alternative="greater")
```






###2. Two sample T test, HW5 #6.6
```{r}
stream <- read.csv("data/stream.csv")
```

```{r}
#Part A
#Equal variances will yield Pooled variance T test, whereas unequal will yield Satterthwaite.

#H0: dowmstream - upstream >= -.5
#Ha: downstream - upstream < -.5

t.test(stream$downstream, stream$upstream, 
       alternative="less", mu=-.5, conf.level=.01,
       var.equal=T)


#Part B
shapiro.test(stream$downstream)
shapiro.test(stream$upstream)

layout(mat = matrix(c(1,2), 2, 1),  height = c(3,1)) #matrix(data, nrow, ncol)
par(mar=c(0, 3, 2, 1)) #margins for each plot
hist(stream$downstream, xlim=c(3,5), col="pink", main="Distribution of Downstream")
boxplot(stream$downstream, horizontal=T, outline=T, frame=F, col="green1", width=10, ylim=c(3,5), axes=F)

qqnorm(stream$downstream)
qqline(stream$downstream)

layout(mat = matrix(c(1,2), 2, 1),  height = c(3,1))
par(mar=c(0, 3, 2, 1))
hist(stream$upstream, xlim=c(4,6), col="pink", main="Distribution of Upstream")
boxplot(stream$upstream, horizontal=T, outline=T, frame=F, col="green1", width=10, ylim=c(4,6), axes=F)

qqnorm(stream$upstream)
qqline(stream$upstream)
```





###3. Paired T test, HW5 #6.28
```{r}
repairpersons <- read.csv("data/ex6-28.txt")
```

```{r}
#Part A
#H0: After - Before >= 0
#Ha: After - Before < 0

t.test(repairpersons$X.After., repairpersons$X.Before., 
       alternative="less", paired=T)


#Part B
#95% Confidence interval

t.test(repairpersons$X.After., repairpersons$X.Before., paired=T)


#Part C
repairpersons <- mutate(repairpersons, diff = X.After. - X.Before.)
shapiro.test(repairpersons$diff)

layout(mat = matrix(c(1,2), 2, 1),  height = c(3,1))
par(mar=c(0, 3, 2, 1))
hist(repairpersons$diff, xlim=c(-17,19), col="pink", main="Distribution of After-Before")
boxplot(repairpersons$diff, horizontal=T, outline=T, frame=F, col="green1", width=10, ylim=c(-17,19), axes=F)

qqnorm(repairpersons$diff)
qqline(repairpersons$diff)
```




###4. Wilcoxon Rank Sum / Mann Whitney U non-parametric test for 2 independent means, Week 7 lab
See Note 4 for processing hairy strings
```{r}
golf <- read.csv("data/golf.csv")
```

```{r}
#The base fn wilcox.test cannot handle ties, so another package is necessary.
library(exactRankTests)

#H0: medians/distributions are the same
#Ha: medians/distributions are different
wilcox.exact(score ~ gender, data=golf)
```




###5. Wilcoxon Signed Rank np test for paired continuous data, Week 7 lab
```{r}
mining <- read.csv("data/mining.csv")
```

```{r}
wilcox.exact(mining$after, mining$before, paired=T)

#The value V = 89 corresponds to the sum of ranks assigned to the differences with positive sign. See Note 5
```





###6. One Way ANOVA, Week 9 lab
See Note 6 for references
```{r}
oil <- read.csv("data/oil.csv")
```

Using car package
```{r}
library(car) #easier to calculate type 3 SS using this package than base R

Anova(lm(visc ~ type, data=oil), type=3)

#HOV Levene test. SAS does Levene's a different way, see Note 6.
leveneTest(visc ~ type, data=oil)
```






###7. Kruskal Wallis np 1 way test, Week 9 lab
```{r}
kruskal.test(visc ~ type, data = oil) 
```




###8. Chi Square test for independence, week 9 lab
```{r}
de <- read.csv("data/de.csv")
```

```{r}
#Chi-square function requires data to be in a table
tbl <- table(de$exposure, de$disease)

#Chi Sq Test for Independence
chisq.test(tbl, correct=F) #Default of correct=T yields Continuity Adj Chi Sq

#Contingency Table
library(gmodels)
CrossTable(de$exposure, de$disease)

#Fisher Exact Test
fisher.test(de$exposure, de$disease)
```




###9. Chi Square goodness of fit test, Week 10 lab
```{r}
gator <- read.csv("data/gator.csv")

#data must be summarized for chisq test
gator_count <- gator %>% 
  group_by(type) %>% 
  summarise(count = n())
```

```{r}
#chisq.test(x = observed, p = expected)
#F I O
chisq.test(x=gator_count$count, p=c(1/2, 3/10, 1/5))
```





###10. Linear Contrasts, Week 3 lab (first lab)
See Note 10 for reference
```{r}
scores <- read.csv("data/scores.csv")

#The order of the contrast weights maps to the order of the levels. Here, I set the order to be alphabetical.
levels(scores$method) <- c("computer", "control", "instructor", "piano")
```



```{r}
#Calculate inverse matrix of contrasts
c1 <- c(1, 0, 0, -1) #computer vs piano
c2 <- c(0, 1, -1, 0) #control vs instructor
c3 <- c(-1, 3, -1,-1) #control vs all

mat_temp <- rbind(constant=1/4, c1, c2, c3)
mat <- solve(mat_temp)

mat <- mat[, -1]

#Calculate contrasts.
model <- lm(score ~ method, data=scores, contrasts=list(method = mat))
summary(model)
```


###11. Tukey, Week 3 lab (first lab, same as above)
Different package used, see Note 11
```{r}
library(TukeyC)
#Tukey won't take lm or Anova, only aov.
fit <- aov(score ~ method, data=scores)

#This package matches SAS output. TukeyC(x, which)
tukes <- TukeyC(fit, "method")
tukes
plot(tukes)
```


###12. Two Way ANOVA, Week 4 lab (second lab)
See Note 12
```{r}
soil <- read.csv("data/soil.csv")

soil$calcium <- as.factor(soil$calcium)
soil$ph <- as.factor(soil$ph)
```
Using car package
```{r}
#Two way Anova
soil_fit <- lm(diameter ~ calcium * ph, data=soil)
Anova(soil_fit, type=3)

# Residuals test for normality
shapiro.test(soil_fit$residuals)

layout(mat = matrix(c(1,2), 2, 1),  height = c(3,1))
par(mar=c(0, 3, 2, 1))
hist(soil_fit$residuals, xlim=c(-1,1), col="pink", main="Distribution of Residuals")
boxplot(soil_fit$residuals, horizontal=T, outline=T, frame=F, col="green1", width=10, ylim=c(-1,1), axes=F)

qqnorm(soil_fit$residuals)
qqline(soil_fit$residuals)
```


Another example from HW4 #14-8

```{r}
#Another example from HW4 #14-8
attention <- read.csv("data/attention.csv")

#Two way anova
Anova(lm(span ~ age * product, data=attention), type=3)
```




###13. Completely Randomized Block Design, HW3 #15-6
```{r}
music <- read.csv("data/music.csv")

music$subject <- as.factor(music$subject)
```

```{r}
Anova(lm(typing_efficiency ~ type_music + subject, data=music), type=3)
```





###14. Latin Square Design (LSD), HW3 #15-40
```{r}
lights <- read.csv("data/lights.csv")

lights$intersection <- as.factor(lights$intersection)
lights$period <- as.factor(lights$period)
```

```{r}
Anova(lm(minutes ~ signaling + intersection + period, data=lights), type=3)

#Bonus tukey
fit <- aov(minutes ~ signaling + intersection + period, data=lights)
tukes <- TukeyC(fit, "signaling")
tukes
plot(tukes)
```








###15. ANCOVA, Week 5 lab
```{r}
oysters <- read.csv("data/oysters.csv")

oysters$treatment <- as.factor(oysters$treatment)
oysters$replication <- as.factor(oysters$replication)
```

```{r}
#ANCOVA
Anova(lm(final_weight ~ initial_weight + treatment, data=oysters), type=3)

#Test for heteroskedasticity (interaction term)
Anova(lm(final_weight ~ initial_weight * treatment, data=oysters), type=3)
```




###16. Random Effect and Mixed Effect Models, HW5 #17-10
See Note 16 for all the testing

```{r}
pesticide <- read.csv("data/pesticide.csv")

pesticide$location <- as.factor(pesticide$location)
pesticide$chemical <- as.factor(pesticide$chemical)
```

```{r}
pest_fit <- lm(number_killed ~ location * chemical, data=pesticide)
Anova(pest_fit, type=3)
```

```{r}
# When you have random effect, divide the individual effects by interaction term.
# Chemical is random and interacts with Location, so calculating the F-stat for both involves dividing by mean square of the interaction, not the error.

# F-value for Location = ms_location/ms_location:chemical
(3.81/4)/(16.16/12) 
pf((3.81/4)/(16.16/12), 4, 12, lower.tail=F)

# F-value for Chemical = ms_chemical/ms_location:chemical
(180.13/3)/(16.16/12)
pf((180.13/3)/(16.16/12), 3, 12, lower.tail=F)
```


Varcomp, see Note 16 for reference.
```{r}
library(VCA)

#Look at VC column in fit$aov.tab
fit <- remlMM(number_killed~(location)*chemical, pesticide, cov=T)
print(fit)
```





###17. Nested Factors, Week 8 lab #1
```{r}
drug <- read.csv("data/drug.csv")

drug$site <- as.factor(drug$site)
drug$batch <- as.factor(drug$batch)
drug$tablet <- as.factor(drug$tablet)
```

```{r}
# Specifying nested term: site/batch == batch(site) in SAS
drug_fit <- Anova(lm(y ~ site + site/batch, data=drug), type=3)
drug_fit
```


```{r}
# Batch is a random effect
# F-stat for site = MS_site / MS_site:batch
# Anova() rounds
0.02/(0.45/4)

# But if you grab from fit object, you can get exact answers
x <- (drug_fit$`Sum Sq`[2]/drug_fit$Df[2])/(drug_fit$`Sum Sq`[3]/drug_fit$Df[3])
x

pf(x, 1, 4, lower.tail=F)
```

Week 8 lab, #2
```{r}
engi <- read.csv("data/engi.csv")

engi$device <- as.factor(engi$device)
engi$rep <- as.factor(engi$rep)
engi$facility <- as.factor(engi$facility)
engi$tester <- as.factor(engi$tester)
```

```{r}
# Tester is nested within facility
engi_fit <- Anova(lm(y ~ facility + device + facility:device + facility/tester, data=engi), type=3)
engi_fit
```

```{r}
# Finding F-stat and p-value for Facility, which has random effect Tester nested within it.
# Approx
(0.935)/(1.106/6)
pf((0.935)/(1.106/6),1,6,lower.tail=F)

# Exact
x <- engi_fit$`Sum Sq`[2]/(engi_fit$`Sum Sq`[5]/engi_fit$Df[5])
x
pf(x, 1, 6, lower.tail=F)
```




###18. Crossover Design, Week 9 lab
```{r}
bp <- read.csv("data/bp.csv")
bp$sequence<- as.factor(bp$sequence)
bp$patient<- as.factor(bp$patient)
bp$period<- as.factor(bp$period)
```

```{r}
# Since interaction is insignificant, rerun model without it.
bp_fit <- Anova(lm(y ~ sequence + treatment + period + sequence/patient, data=bp), type=3)
bp_fit
```

```{r}
# Patient is nested within Sequence, and thus is considered a random effect.
# F-stat for Sequence = MS_seq / MS_seq:pat
(bp_fit$`Sum Sq`[2]/bp_fit$Df[2])/(bp_fit$`Sum Sq`[5]/bp_fit$Df[5])
pf((bp_fit$`Sum Sq`[2]/bp_fit$Df[2])/(bp_fit$`Sum Sq`[5]/bp_fit$Df[5]), 2,9,lower.tail = F)
```








###Note 0
Reference for changing options:
https://www.r-bloggers.com/ensuring-r-generates-the-same-anova-f-values-as-spss/)

###Note 4
```{}
#Use gsub for messier strings. Here it's unnecessary.
#golf <- read.table(text=gsub("(?<=[a-z])\\s+", " ", golf_text, perl=TRUE), header=FALSE, col.names = c("gender", "score"))
```

###Note 5
```{}
#Calculation of V:
diff <- mining$after - mining$before
diff <- diff[diff!=0] #delete all differences = 0, as per Signed Rank procedure
diff.rank <- rank(abs(diff))
diff.rank.sign <- diff.rank * sign(diff[])
ranks.pos <- sum(diff.rank.sign[diff.rank.sign > 0]) 
ranks.neg <- -sum(diff.rank.sign[diff.rank.sign < 0]) 
```

###Note 6
Referenced used for ANOVA:
https://www.r-bloggers.com/anova-%E2%80%93-type-iiiiii-ss-explained/
```{}
#SAS calculates Leven using Squared deviations. R and it's packages seem to do Absolute instead.
#The following replicates what SAS does.
#https://stackoverflow.com/questions/22396543/levenes-test-using-the-squared-residuals

oil.lm <- lm(visc ~ type, data = oil)
anova(lm(residuals(oil.lm)^2 ~ oil$type))

#HOV Levene using absolute deviations
leveneTest(visc ~ type, data=oil)
```
Using ezANOVA package
```{}
library(ez)

#ezANOVA requires an ID column, wid. This is so that it can handle repeated measures.
oil$id <- as.factor(c(1:length(oil$type)))


av <- ezANOVA(data=oil, dv=visc, wid=id ,between=type, detail=T, type=3, white.adjust=T)
print(av)
```
ANT example
```{}
data(ANT)
b_anova_full <- ezANOVA(data=ANT,
                        dv=rt,
                        wid=subnum,
                        between=group,
                        within_full=.(cue, flank),
                        type=3,
                        detailed=T,
                        return_aov=T)
print(b_anova_full)
```

###Note 10
Reference, see DIY Contrasts:
https://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html

###Note 11
```{}
#Base R uses TukeyHSD, which yields less elegant output.
tukes <- TukeyHSD(fit, "method")
tukes
plot(tukes)
```

###Note 12
Using ezANOVA package
```{}
soil$id <- as.factor(c(1:length(soil$diameter))) #ezANOVA requires an id column
soil_anova <- ezANOVA(data=soil, dv=diameter, wid=id, between=.(calcium, ph), type=3, detailed=T, return_aov=T)

resids <- resid(soil_anova$aov) #bc we used return_aov=T, we can call the base aov() and output the residuals

shapiro.test(resids)

layout(mat = matrix(c(1,2), 2, 1),  height = c(3,1))
par(mar=c(0, 3, 2, 1))
hist(resids, xlim=c(-1,1), col="pink", main="Distribution of Residuals")
boxplot(resids, horizontal=T, outline=T, frame=F, col="green1", width=10, ylim=c(-1,1), axes=F)

qqnorm(resids)
qqline(resids)
```

###Note 16
reference: https://mcfromnz.wordpress.com/2011/03/02/anova-type-iiiiii-ss-explained/
Varcomp reference: https://stats.stackexchange.com/questions/161225/estimates-of-the-variance-of-the-variance-component-of-a-mixed-effects-model
Testing on paint
```{}
paint <- read.csv("data/CH17/ex17-1.txt")
colnames(paint) <- c("batch", "percentage")
```
Testing on pesticide
```{}
#Both of these attempts have chemical F = 48.11, but answer should be 173.41 if divided by ms_error, or 44.49 if divided by ms_interaction

#attempt 1 with interaction, wrong
library(afex)
mixed(number_killed ~ chemical + (1|location) + (1|location:chemical), data=pesticide)

#with no interaction, CORRECT, but that means interaction is wrong...
model2 <- lmer_alt(number_killed ~ chemical + (1|location), data=pesticide)
anova(model2)

#attempt 2 with interaction, wrong
model3 <- lmer_alt(number_killed ~ chemical + (1|location) + (1|chemical:location), data=pesticide)
summary(model3)
```
Working solution (but answer is type 2)
```{}
#Make F table manually
fit <- lm(number_killed ~ location * chemical, data=pesticide)
results <- Anova(fit) #this is the anova table. The following replaces F tests with MS_main/MS_interaction
Df <- results$Df
SumSq <- results$"Sum Sq"
MeanSq <- SumSq/results$Df
Fvalue <- results$"F value"
Pvalue <- results$"Pr(>F)"
Error.Term <- MeanSq[3]
df.error <- Df[3]

Fvalue[1] <- MeanSq[1]/Error.Term
Pvalue[1] <- 1 - pf(Fvalue[1], Df[1], df.error)

Fvalue[2] <- MeanSq[2]/Error.Term
Pvalue[2] <- 1 - pf(Fvalue[2], Df[2], df.error)

Ftable <- cbind(Df, SumSq, MeanSq, Fvalue, Pvalue)
rownames(Ftable) <- c("Locations", "Chemicals", "Locations:Chemicals", "Residuals")
print(Ftable)
```
Misc things
```{}
library(lmerTest) 

coef(fit)


#other attempts
model <- lmerTest::lmer(number_killed ~ chemical + (1|location) + (1|location:chemical), data=pesticide)
rand(model)

linearHypothesis(model, "chemical=0") #doesnt work
```

###Note 18
For carryover, unused and unfinished
```{}
carryover <- trt_str
carryover$period <- as.factor(as.numeric(trt_str$period) + 1)
```
Regarding multicolinearity:
https://stats.stackexchange.com/questions/112442/what-are-aliased-coefficients